---
title: "Assignment 1"
author: "Fernando_Almeida"
date: "2024-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Prediction Assignment

## Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: 
http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har
 (see the section on the Weight Lifting Exercise Dataset).

## Data
The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


The data for this project come from this source: 
http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har
. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.

```{r packages}
# Installing required packages
install.packages("caret", repos='http://cran.us.r-project.org')
library(caret)

install.packages("randomForest", repos='http://cran.us.r-project.org')
library(randomForest)

install.packages("rpart", repos='http://cran.us.r-project.org')
library(rpart)
```

## Data download
At first data is downloaded and training data splited into training and testing subsets.

```{r data_download}
# Set seed for reproducibiity
set.seed(1)

# Download data
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"   
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv" 

# Cleaning NAs, #DIV/0! and empty spaces
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))  
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))

#Remove variables with near zero variance
training<-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]

#Remove columns that are not predictors
training   <-training[,-c(1:7)]
testing <-testing[,-c(1:7)]

training$classe <- as.factor(training$classe)

# Split training data in training (70%) and testing (30%) subsets:
inTrain <- createDataPartition(y=training$classe, p=0.7, list=FALSE)    
newTraining <- training[inTrain, ]
newValidating <- training[-inTrain, ]
```

## Decision Tree
The first model used was Decision Tree, which presented accuracy about 75%.

``` {r decision_tree}
# DECISION TREE - Fit model in newTraining 
fitDT <- rpart(classe ~ ., data=newTraining, method="class")

# Use model to predict class in validation
predictionDT <- predict(fitDT, newValidating, type = "class")

# Estimating errors in Decision Tree Model
confusionMatrix(newValidating$classe, predictionDT)
```

## Random Forest
The second prediction model was Random Forest, which presented accuracy higher than 95%.

``` {r random_forest}
# RANDOM FOREST - Fit model in newTraining 
fitRF <- randomForest(classe ~ ., data=newTraining, method="class")

# Use model to predict class in validation
predictionRF <- predict(fitRF, newValidating, type = "class")

# Estimating errors in Random Forest Model
confusionMatrix(newValidating$classe, predictionRF)
```

## Submission Prediction
As random forest model presented higher accuracy, it was used to predict testing data.

```{r submission_prediction}
# As random forest presented higher accuracy, it will be used to predict submission
predictSubmission <- predict(fitRF, testing, type="class")
predictSubmission
```